Future challenges for Research Software Engineering

### 2025-05-19 future_challenges_for_RSEng.tex
Extend and structure the introduction (#38)
* Add references to previous work

* Introduction: Replace reference with footnote, split paragraphs

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are common in industry \cite{gartner}, so far there is a projection lacking for the RSEng space.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\cite{Goth2025EndRSEng}, without
claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how Research Software Engineering should, or could, develop.
It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}.
It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.
Nevertheless, in order to structure the discussion, we decided to consider
the evolution of the term RSE and its definition,
the growth and development of the RSE community,
the impact of automation and generative artificial intelligence in RSE work,
the impact of complexity and usability in research software,
ethical and social considerations,
the expected transformations in academic institutions,
and the potential impact of alternative career paths for RSEs and new groups of RTPs (\emph{Research Technical Professional}s).
%
We conclude the paper with a summary and with our wishes and recommendations.

>>>>

Research Software Engineering is still a developing field,
with increasing integration into various aspects of research.
In Germany, a formal call for creating a sustainable environment for research software
is only a couple of years behind~\cite{Anzt2021},
while the foundational competencies of a research software engineer have only recently been defined in detail~\cite{Goth2024}. Central RSE units are in the process of being imagined~\cite{Kempf2025-draft},
while an RSE Master's is in the early stages of design.

In this opinion paper, we try to imagine the state of our field
and community in the mid-term future, focusing on future challenges.
While future projections are common in industry \cite{gartner}, so far there is a projection lacking for the RSEng space.
We present here the results of a workshop discussion at the
deRSE25 conference in Karlsruhe\footnote{deRSE25, session ``The End of {RSEng}? Challenges and Risks for {RSEng}'': \url{https://events.hifis.net/event/1741/contributions/14026/}}, without
claiming to represent the entire deRSE community.
Nevertheless, by disseminating the discussed ideas,
we aim to engage the community in creating a future worth living
and to advance the conversation about how Research Software Engineering should, or could, develop.
It is already difficult to structure the spectrum of RSE tasks and responsibilities today given the spectrum of research software \cite{hasselbring2024}.
It is even more challenging to structure the aspects that will become relevant for the RSE community in the future.

In order to structure the discussion, we decided to consider
the evolution of the term RSE and its definition,
the growth and development of the RSE community,
the impact of automation and generative artificial intelligence in RSE work,
the impact of complexity and usability in research software,
ethical and social considerations,
the expected transformations in academic institutions,
and the potential impact of alternative career paths for RSEs and new groups of RTPs (\emph{Research Technical Professional}s).
%
We conclude the paper with a summary and with our wishes and recommendations.
